Vagueness, Noise, and Signaling

### 2015-05-13 paper/version_03/paper.tex
Some more minor tweaks. Added descriptions for definitions of diffusion. Labelling within group measure figure using subfigures.

The replicator diffusion dynamic is a noise-perturbed variant of the
\rd in behavioral strategies. Fix a sim-max game with $\States =
\Acts$ and a confusion matrix $C : \States \times \States$. $C$ is a
row-stochastic matrix, i.e., each row is a probability
distribution. Elements $C_{ij}$ give the probability that $\state_i$
is realized as $\state_j$. The confusability of states affects senders
and receivers alike, but in slightly different ways (see
Figure~\ref{fig:noise-perturbation-of-strategies}). For the sender,
$C_{ij}$ is the probability that the actual state \mystate{i} is
perceived as \mystate{j}. For the receiver, $C_{ij}$ is the
probability that \mystate{j} is the interpretation that is actually
formed when \mystate{i} is the intended interpretation.

@@

The aggregate effect of confusion of states on behavioral strategies
can be captured in a function that maps behavioral strategies
$\Sstrat$ and $\Rstrat$ to their \emph{diffusions} $D(\Sstrat)$ and
$D(\Rstrat)$. Suppose that $\Sstrat$ and $\Rstrat$ represent a
population's average signaling behavior if there was no confusion of
states. Then $D(\Sstrat)$ and $D(\Rstrat)$ give the expected aggregate
population behavior under the actual noise-perturbed behavior of each
agent. Since the effect of confusion of states is that behavior at one
choice point percolates to behavior at similar choice points, we speak
of \emph{diffusion of behavior under confusion of states}. If we
conceive of behavioral strategies $\Sstrat$ and $\Rstrat$ as
row-stochastic matrices (rows are choice points, columns are choices;
each row is a probability distribution), the diffusion effect of
confusion matrix $C$ is just matrix multiplication:
\begin{align}
  \label{eq:confusion-function}
  \Diff_C(\Sstrat) & = C \Sstrat &    \Diff_C(\Rstrat) & = \Rstrat C\,.
\end{align}

@@

The sequential definition highlights that other (discrete-time)
evolutionary dynamics can be subjected to state-confusability in an
analogous fashion. It is easy, for instance, to perturb a
best-response dynamic in case we wanted to model cognizing, rational
agents with a noisy perceptual system. Our choice of the \rd as a
carrier dynamic is motivated by its generality and austerity, being
compatible with a variety of implicit agent-level behavior, without
presupposing extravagant higher cognitive abilities
\citep[c.f.][]{Sandholm2013:Population-Game}. Moreover, by using the
\rd as a carrier for diffusion, we actually obtain a special case of a
behavioral-strategy version of the replicator mutator dynamics. This
is important to know because it enables similar applications of
diffusion from perceptual confusion also in models of biological
evolution.

%The following section works out the relation between the \rdd and the \rmd.
%Although this section is conceptually important, it is also rather
%technical and not strictly necessary to appreciate the main point of
%the paper.

@@

Confusion of states should intuitively be a function of their
perceptual similarity. To make this concrete, let us assume that the
state space of the sim-max game consists of $n \ge 2$ states that are
equally spaced across the unit interval, including $0$ and $1$. The
distance $\card{\mystate{i} - \mystate{j}}$ is the objective, physical
similarity between two states $\mystate{i}$ and
$\mystate{j}$. Distance in physical space feeds into a perceptual
similarity function, as described by
\citet{Nosofsky1986:Attention-Simil}:
\begin{align*}
  \similarity(\mystate{i}, \mystate{j} \myts ; \myts \impairment) =
      \begin{cases}
    1 & \textrm{if } \impairment = 0 \textrm{ and } \mystate{i} = \mystate{j} \\
    0 & \textrm{if } \impairment = 0 \textrm{ and } \mystate{i} \neq \mystate{j} \\
 \expo \left ( -  \frac{\card{\mystate{i} - \mystate{j}}^s}{ \impairment^2} \right ) & \textrm{otherwise,} \\
    \end{cases}
\end{align*}
where $\impairment \ge 0$ is an impairment, imprecision or indiscriminability
parameter\footnote{We use these terms interchangeably throughout the paper.}.
When $\impairment=0$ agents perfectly discriminate between
states; when $\impairment \rightarrow \infty$ agents cannot
discriminate states at all. Figure~\ref{fig:NosofskySim} gives an
impression of Nosofsky-similarity for different parameter
values. Other formalizations of perceptual similarity are possible,
including ones that allow for different discriminability in different
areas of the state space, but we stick with Nosofsky's similarity
function for the time being, because it is mathematically simple, yet
an established notion in mathematical psychology.

@@

We further assume that the probability of confusing any two states
$\mystate{i}$ and $\mystate{j}$ is proportional to their perceived
similarity and therefore obtained by normalization:
\begin{align*}
  C_{ij} = \frac{\similarity(\mystate{i}, \mystate{j} \myts ; \myts
  \impairment)} {\sum_j \similarity(\mystate{i}, \mystate{j} \myts ; \myts
  \impairment)} \,.
\end{align*}
Confusion of states is then also a function of imprecision parameter
$\impairment$. For $\impairment = 0$, the confusion matrix has $C_{ii}
= 1$; for $\impairment > 0$, $C$ is positive, i.e., $C_{ij} >0$ for
all $i$ and $j$; for $\impairment \rightarrow \infty$, we find $C_{ij}
= \nicefrac{1}{\card{\States}}$ for all $i$ and $j$.

@@

Representative examples for resulting strategy pairs are given in
Figure~\ref{fig:example_strats}. Figure~\ref{fig:example_stratsA}
shows a strategy pair at stopping time with 90 states, tolerance
$\toler = 0.1$ and imprecision $\impairment = 0$. Zero imprecision
means that the trial was effectively an application of the plain
\rd. Noteworthily, the given sender strategy approximates a pure
sender strategy that crisply partitions the state space into
non-convex sets. The irregular shape of the receiver strategy shows
clearly that the pictured strategy pair has not yet reached a
dynamically stable state under the \rd. Indeed, the trial was stopped
after the maximum of 200 rounds. In contrast, the outcome of a trial
with identical parameters, except with imprecision $\impairment =
0.1$, which is shown in Figure~\ref{fig:example_stratsB}, had
converged (in our technical sense) after 99 rounds of iterated
\rdd. The sender strategy shows a smooth blending from one
``category'' to the other, and the receiver's interpretations are
rather extended curves, peaking at a central point in the relevant
``categories.''

@@

Figure~\ref{fig:MeanMetrics} shows plots summarizing the results of
our simulation. For perspicuity, we only plot results for one level of
tolerance $\beta = 0.1$, and leave out the case of $n_s = 90$. Still,
every qualitative trend we mention in the following applies to the
whole set of results.

@@

Interestingly, this type of variability in evolutionary outcomes can
be weeded out by impairment. To investigate this, we calculated the average
distance between evolved sender strategies within each group of trials
that had identical parameter values. We determined the distance
between strategies $\Sstrat$ and $\Sstrat'$ as the average Hellinger
distance between distributions $\Sstrat(\state)$ and
$\Sstrat'(\state)$ at each choice point $\state$:
\begin{align*}
  \text{HD}(\Sstrat,\Sstrat') & = \frac{1}{{\card{\States} \cdot
     \sqrt{2}}} \cdot  \sum_{\state \in \States} 
 \sqrt{\sum_{\messg \in  \Messgs}
         \left ( \sqrt{\Sstrat(\state,\messg)} -
         \sqrt{\Sstrat'(\state,\messg)} \right )^2 }\,.
\end{align*}
To compensate for the arbitrariness of message use, we set the
distance between strategies $\Sstrat$ and $\Sstrat'$ to be the maximum
of $\text{HD}(\Sstrat,\Sstrat')$ and $\text{HD}(\Sstrat^*,\Sstrat')$
where $\Sstrat^*$ is $\Sstrat$ with reversed message indices. An example of the
\emph{within group distance}, i.e., the average distances between all
sender strategies obtained for the same parameter values, is plotted
on the left in Figure~\ref{fig:GroupMeasures} for $\beta =
0.1$ and $n_s = 10$. Despite some quantitative differences, the
general trend is the same for all other parameter settings that we tested:
with increasing imprecision, the sender strategies evolving under \rdd
were much more alike (modulo swapping of messages). This means that
perceptual imprecision unifies evolutionary outcomes and guarantees
sender strategies that are not only convex, but also regular in that
they induce a vague category split exactly in the middle of the unit
interval.

@@

    \includegraphics[width=0.475\textwidth]{plots/WithinGroupDistanceConcise.pdf}
    \hfill
    \includegraphics[width=0.475\textwidth]{plots/WithinGroupEUConcise.pdf}

    \caption{Within group distance (left) and within group expected
      utility (right) as a function of impairment for all runs with
      $\toler = 0.1$ and $n_s = 10$.}
  \label{fig:GroupMeasures}
\end{figure}

This unifying property of perceptual imprecision could be considered
an evolutionary beneficial side-effect. Higher imprecision can lead to
higher \emph{within group expected utility}, defined as the average
expected utility that each evolved language (i.e., pair of sender and
receiver strategy at stopping time) scored when playing against an
arbitrary other language obtained for the same parameter values. The
right side of Figure~\ref{fig:GroupMeasures} gives a representative
example. The observation repeats for other parameter values: while
imprecision might decrease the communicative efficiency of individual
languages, it increases the conceptual coherence and communicative
success between independently evolving strategies. It is almost as if
mere confusability of states imposes a regularity constraint on
evolving categories.

>>>>

The replicator diffusion dynamic is a noise-perturbed variant of the
\rd in behavioral strategies. Fix a sim-max game with $\States =
\Acts$ and a confusion matrix $C : \States \times \States$, where $C$ is
row-stochastic, i.e., each row is a probability
distribution. Elements $C_{ij}$ give the probability that $\state_i$
is realized as $\state_j$. The confusability of states affects senders
and receivers alike, but in slightly different ways. For the sender,
$C_{ij}$ is the probability that the actual state \mystate{i} is
perceived as \mystate{j}. For the receiver, $C_{ij}$ is the
probability that \mystate{j} is the interpretation that is actually
formed when \mystate{i} is the intended interpretation.
See Figure~\ref{fig:noise-perturbation-of-strategies} for an illustration.

@@

The aggregate effect of confusion of states on behavioral strategies
can be captured in a function that maps behavioral strategies
$\Sstrat$ and $\Rstrat$ to their \emph{diffusions} $D(\Sstrat)$ and
$D(\Rstrat)$. Suppose that $\Sstrat$ and $\Rstrat$ represent a
population's average signaling behavior if there was no confusion of
states. Then $D(\Sstrat)$ and $D(\Rstrat)$ give the expected aggregate
population behavior under the actual noise-perturbed behavior of each
agent. Since the effect of confusion of states is that behavior at one
choice point percolates to behavior at similar choice points, we speak
of \emph{diffusion of behavior under confusion of states}. If we
conceive of behavioral strategies $\Sstrat$ and $\Rstrat$ as
row-stochastic matrices (rows are choice points, columns are choices;
each row is a probability distribution), we can define the diffusion effect of
confusion matrix $C$ as matrix multiplication:
\begin{align}
  \label{eq:confusion-function}
  \Diff_C(\Sstrat) & = C \Sstrat &    \Diff_C(\Rstrat) & = \Rstrat C\,.
\end{align}
For example, diffusion thus defined determines that, on the diffused sender strategy, the probability that message $\messg$ is chosen given a state $\state_i$ is given by $\sum_{j} C_{ij} \cdot \Sstrat(\messg \probbar \state_j)$.
This means that the original probability given by $\Sstrat$ is increased by the original probability of this message being chosen in other states, proportionally to how likely it is that they are mistaken for $\state_i$.

@@

The sequential definition highlights that other (discrete-time)
evolutionary dynamics can be subjected to state-confusability in an
analogous fashion. It is easy, for instance, to perturb a
best-response dynamic in case we wanted to model cognizing, rational
agents with a noisy perceptual system. Our choice of the \rd as a
carrier dynamic is motivated by its generality and austerity, being
compatible with a variety of implicit agent-level behavior, without
presupposing extravagant higher cognitive abilities
\citep[c.f.][]{Sandholm2013:Population-Game}. Moreover, by using the
\rd as a carrier for diffusion, we actually obtain a special case of a
behavioral-strategy version of the replicator mutator dynamics. This
is important to know because it enables similar applications of
diffusion from perceptual confusion also in models of biological
evolution.
We worked out the relation between the \rdd and the \rmd, but given that it is rather technical and not strictly necessary to appreciate the main point of the paper, we defer it to Appendix~\ref{sec:diffusion-as-special}.
In the following section we explore the long-term effects of using the \rdd.

@@

Confusion of states should intuitively be a function of their
perceptual similarity: the more similar two states are, the more likely they could be mistaken for each other.
To make this concrete, let us assume that the
state space of the sim-max game consists of $n \ge 2$ states that are
equally spaced across the unit interval, including $0$ and $1$. The
distance $\card{\mystate{i} - \mystate{j}}$ is the objective, physical
similarity between two states $\mystate{i}$ and
$\mystate{j}$. Distance in physical space feeds into a perceptual
similarity function, as described by
\citet{Nosofsky1986:Attention-Simil}:
\begin{align*}
  \similarity(\mystate{i}, \mystate{j} \myts ; \myts \impairment) =
      \begin{cases}
    1 & \textrm{if } \impairment = 0 \textrm{ and } \mystate{i} = \mystate{j} \\
    0 & \textrm{if } \impairment = 0 \textrm{ and } \mystate{i} \neq \mystate{j} \\
 \expo \left ( -  \frac{\card{\mystate{i} - \mystate{j}}^s}{ \impairment^2} \right ) & \textrm{otherwise,} \\
    \end{cases}
\end{align*}
where $\impairment \ge 0$ is an impairment, imprecision or indiscriminability
parameter\footnote{We use these terms interchangeably throughout the paper.}.
When $\impairment=0$ agents perfectly discriminate between
states; when $\impairment \rightarrow \infty$ agents cannot
discriminate states at all. Figure~\ref{fig:NosofskySim} gives an
impression of Nosofsky-similarity for different parameter
values. Other formalizations of perceptual similarity are possible,
including ones that allow for different discriminability in different
areas of the state space, but we stick with Nosofsky's similarity
function for the time being, because it is mathematically simple, yet
an established notion in mathematical psychology.

@@

We further assume that the probability of confusing any two states
$\mystate{i}$ and $\mystate{j}$ is proportional to their perceived
similarity and therefore obtained by normalization:
\begin{align*}
  C_{ij} = \frac{\similarity(\mystate{i}, \mystate{j} \myts ; \myts
  \impairment)} {\sum_j \similarity(\mystate{i}, \mystate{j} \myts ; \myts
  \impairment)} \,.
\end{align*}
Confusion of states is then also a function of imprecision parameter
$\impairment$. For $\impairment = 0$, the confusion matrix has $C_{ii}
= 1$ and $C_{ij} = 0$ for $i \neq j$. For $\impairment > 0$, $C$ is positive, i.e., $C_{ij} >0$ for
all $i$ and $j$. For $\impairment \rightarrow \infty$, we find $C_{ij}
\rightarrow \nicefrac{1}{\card{\States}}$ for all $i$ and $j$.

@@

Representative examples for resulting strategy pairs are given in
Figure~\ref{fig:example_strats}. Figure~\ref{fig:example_stratsA}
shows a strategy pair at stopping time with 90 states, tolerance
$\toler = 0.1$ and imprecision $\impairment = 0$. Zero imprecision
means that the trial was effectively an application of the plain
\rd. Noteworthily, the given sender strategy approximates a pure
sender strategy that crisply partitions the state space into
non-convex sets. The irregular shape of the receiver strategy shows
clearly that the pictured strategy pair has not yet reached a
dynamically stable state under the \rd. Indeed, the trial was stopped
for reaching the maximum of 200 rounds. In contrast, the outcome of a trial
with identical parameters, except with imprecision $\impairment =
0.1$, which is shown in Figure~\ref{fig:example_stratsB}, had
converged (in our technical sense) after 99 rounds of iterated
\rdd. The sender strategy shows a smooth blending from one
``category'' to the other, and the receiver's interpretations are
rather extended curves, peaking at a central point in the relevant
``categories.''

@@

Figure~\ref{fig:MeanMetrics} shows plots summarizing some results.
%of our simulation.
For perspicuity, we only plot results for one level of
tolerance $\beta = 0.1$, and leave out the case of $n_s = 90$. Still,
every qualitative trend mentioned in the following applies to the
whole set of results.

@@

Interestingly, this type of variability in evolutionary outcomes can
be weeded out by impairment. To investigate this, we calculated the average
distance between evolved sender strategies within each group of trials
that had identical parameter values. We determined the distance
between strategies $\Sstrat$ and $\Sstrat'$ as the average Hellinger
distance between distributions $\Sstrat(\state)$ and
$\Sstrat'(\state)$ at each choice point $\state$:
\begin{align*}
  \text{HD}(\Sstrat,\Sstrat') & = \frac{1}{{\card{\States} \cdot
     \sqrt{2}}} \cdot  \sum_{\state \in \States} 
 \sqrt{\sum_{\messg \in  \Messgs}
         \left ( \sqrt{\Sstrat(\state,\messg)} -
         \sqrt{\Sstrat'(\state,\messg)} \right )^2 }\,.
\end{align*}
To compensate for the arbitrariness of message use, we set the
distance between strategies $\Sstrat$ and $\Sstrat'$ to be the maximum
of $\text{HD}(\Sstrat,\Sstrat')$ and $\text{HD}(\Sstrat^*,\Sstrat')$
where $\Sstrat^*$ is $\Sstrat$ with reversed message indices. An example of the
\emph{within group distance}, i.e., the average distances between all
sender strategies obtained for the same parameter values, is plotted in Figure~\ref{fig:GroupMeasuresA} for $\beta =
0.1$ and $n_s = 10$. Despite some quantitative differences, the
general trend is the same for all other parameter settings that we tested:
with increasing imprecision, the sender strategies evolving under \rdd
were much more alike (modulo swapping of messages). This means that
perceptual imprecision unifies evolutionary outcomes and guarantees
sender strategies that are not only convex, but also regular in that
they induce a vague category split exactly in the middle of the unit
interval.

@@

  \begin{subfigure}[]{0.45\textwidth}
    \includegraphics[width=\textwidth]{plots/WithinGroupDistanceConcise.pdf}
    \caption{Within group distance}
    \label{fig:GroupMeasuresA}
  \end{subfigure}
    \hfill
  \begin{subfigure}[]{0.45\textwidth}
    \includegraphics[width=\textwidth]{plots/WithinGroupEUConcise.pdf}
    \caption{Within group expected utility}
    \label{fig:GroupMeasuresB}
  \end{subfigure}

    \caption{Within group measures for all runs with $\toler = 0.1$ and $n_s = 10$.}
  \label{fig:GroupMeasures}
\end{figure}

This unifying property of perceptual imprecision could be considered
an evolutionary beneficial side-effect. Higher imprecision can lead to
higher \emph{within group expected utility}, defined as the average
expected utility that each evolved language scored when playing against an
arbitrary other language obtained for the same parameter values. Figure~\ref{fig:GroupMeasuresB} gives a representative
example. The observation repeats for other parameter values: while
imprecision might decrease the communicative efficiency of individual
languages, it increases the conceptual coherence and communicative
success between independently evolving strategies. It is almost as if
mere confusability of states imposes a regularity constraint on
evolving categories.
